---
title: "TransmissionRegressionModeling"
author: "Kevin Roche"
date: "22/10/2021"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
    fig_height: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

Completed as the course project for JHU's Regression Models course. The assignment is as follows:

*You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:*

*1. Is an automatic or manual transmission better for MPG?*
*2. Quantify the MPG difference between automatic and manual transmissions.*

You can read more about the mtcars dataset [here.](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html)

# Setup

```{r}
## Load packages
library(tidyverse)
library(ggplot2)
```

# Data Processing

Let's take a quick look at the dataset we'll be using for the analysis, mtcars.

```{r}
## Load dataset
data(mtcars) 

## Take a look at dimensions, etc.
head(mtcars, 3)
dim(mtcars)
map(mtcars, class)
```

The dataset has 11 columns and 32 rows, a somewhat small sample size. All of the variables are coded as numeric variables, when in reality, quite a few of them are factor variables. Let's fix that.

```{r}
## Change appropriate variables to factor class
mtcars <- mtcars %>% 
        mutate(across(c(cyl, vs, am, gear, carb), ~ as.factor(.)))
```

# Exploratory Data Analysis

Let's plot miles per gallon by transmission type and see if there's a noticeable difference between manual and automatic transmissions.

```{r}
## Miles Per Gallon by Transmission Type
mtcars %>% ggplot(aes(x = am, y = mpg, fill = am)) +
        ggthemes::scale_fill_tableau() +
        geom_boxplot(show.legend = FALSE) +
        geom_point(show.legend = FALSE) +
        labs(x = "Transmission Type", y = "Miles Per Gallon") +
        ggtitle("Miles Per Gallon by Transmission Type") +
        scale_x_discrete(labels = c("Automatic", "Manual"))
        theme_bw() 
```